---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  tidy.opts = list(blank = FALSE)
)
```

# powerjoin

WIP 

{powerjoin} extends {dplyr}'s join functions.

It replaces my {safejoin} package which has un unfortunate homonym on CRAN and
has a suboptimal interface and implementation.

The {dplyr} team is credited since a lot of code is taken from {dplyr} directly.

## Installation

``` r
remotes::install_github("moodymudskipper/powerjoin")
```

## Safe joins

The `check` argument  receives a list provided by `pj_check()`, whose argument
can be :

* NA : stay silent (default except for `implicit_by`)
* "inform"
* "warn"
* "abort"

```{r, error = TRUE}
library(powerjoin)
library(dplyr, warn.conflicts = FALSE)

#-------------------------------------------------------------------------------
# works like {dplyr} by default 
power_left_join(
  band_members, 
  band_instruments)
#-------------------------------------------------------------------------------
# silence the implicit join
power_left_join(
  band_members, 
  band_instruments, 
  check = pj_check(implicit_by = NA))
#-------------------------------------------------------------------------------
# fail unless explicit `by`
power_left_join(
  band_members, 
  band_instruments, 
  check = pj_check(implicit_by = "abort"))
#-------------------------------------------------------------------------------
# fail if column conflict
power_left_join(
  band_members, 
  band_members,
  by = "name",
  check = pj_check(column_conflict = "abort")
)
#-------------------------------------------------------------------------------
# fail if duplicate keys in the left table
power_left_join(
  bind_rows(band_members, band_members[1:2,]), 
  band_instruments,
  by = "name",
  check = pj_check(duplicate_keys_left = "abort")
)
#-------------------------------------------------------------------------------
# fail if unmatched keys in the left table
power_left_join(
  band_members, 
  band_instruments,
  by = "name",
  check = pj_check(unmatched_keys_left = "abort")
)
#-------------------------------------------------------------------------------
# fail if missing keys combination in the left table
power_left_join(
  band_members, 
  band_members,
  check = pj_check(missing_key_combination_left = "abort")
)
#-------------------------------------------------------------------------------
# fail if inconsistent factor levels
power_left_join(
  band_members %>% mutate(name = factor(name)), 
  band_instruments %>% mutate(name = factor(name)), 
  check = pj_check(inconsistent_factor_levels = "abort")
)
#-------------------------------------------------------------------------------
# fail if inconsistent type
power_left_join(
  band_members %>% mutate(name = factor(name)), 
  band_instruments, 
  check = pj_check(inconsistent_type = "abort")
)
```

## Preprocessing inputs

Traditionally key columns need to be repeated when preprocessing inputs 
before a join, we offer a way around this.

compare the following versions of {dplyr} joins vs {powerjoin}

```{r}
library(dplyr, warn.conflicts = FALSE)

#-------------------------------------------------------------------------------
x <- tibble(Species = "setosa", Sepal.Length = 5.1)

# with {dplyr}, select key columns + columns to join
left_join(
  x,
  iris %>% select(Species, Sepal.Length, Petal.Length)
)


#-------------------------------------------------------------------------------
# with {powerjoin}, only select columns to join
power_left_join(
  x,
  iris %>% select_keys_and(Petal.Length)
)

#-------------------------------------------------------------------------------
x <- tibble(Species = c("setosa", "virginica"))

# with {dplyr}, group by key columns first then summarize
left_join(
  x,
  iris %>% group_by(Species) %>% summarize(Sepal.Length = mean(Sepal.Length), .groups = "drop")
)


#-------------------------------------------------------------------------------
# with {powerjoin}, summrize by key columns right away 
power_left_join(
  x,
  iris %>% summarize_by_keys(Sepal.Length = mean(Sepal.Length))
)
```

We have more of these, all variants of tidyverse functions :

* `nest_by_keys()` nests given columns, or all by default, if `name` is given
a sinle list column of data frames is created. 
* `pack_along_keys()` packs given columns, or all non key columns by default, into
a a data frame column named by the `name` argument, it's useful to namespace the
data and avoid conflicts
* `pivot_wider_by_keys()` and `pivot_longer_by_keys()` assume the "id" columns are the keys

## Handle column conflict

To resolve conflicts between identically named join columns, set the `conflict`
argument to a 2 argument function that will take as arguments the 2 conflicting 
joined columns after the join.

coalescing is the most common use case but the feature is flexible

```{r}
df1 <- tibble(id = 1:3, value = c(10, NA, 30))
df2 <- tibble(id = 2:4, value = c(22, 32, 42))

#-------------------------------------------------------------------------------
# no conflict handling
power_left_join(df1, df2, by = "id")

#-------------------------------------------------------------------------------
# coalescing conflicting cols
power_left_join(df1, df2, by = "id", conflict = dplyr::coalesce)


#-------------------------------------------------------------------------------
# we are operating on vectors by default, not row wise!
power_left_join(df1, df2, by = "id", conflict = ~ sum(.x, .y, na.rm = TRUE))


#-------------------------------------------------------------------------------
# for row wise operations, type `rw` on the left side of the formula
power_left_join(df1, df2, by = "id", conflict = rw ~ sum(.x, .y, na.rm = TRUE))
```
